### 2014-05-11 paper/paper.tex
Add an abstract and an implementation

\documentclass[10pt,letter]{article}

\usepackage{amsmath}
\usepackage{amssymb}
% packages that allow mathematical formatting

\usepackage{graphicx}
% package that allows you to include graphics

\usepackage{setspace}
% package that allows you to change spacing

\usepackage{algpseudocode}
\usepackage{algorithmicx}
\usepackage{algorithm}

\usepackage[sc,osf]{mathpazo}
\linespread{1.05}         % Palatino needs more leading (space between lines)
\usepackage[T1]{fontenc}

\fontencoding{T1}
\fontfamily{ppl}
\fontseries{m}
\fontshape{n}
\fontsize{10}{13}
% Set font size. The first parameter is the font size to switch to; the second
% is the \baselineskip to use. The unit of both parameters defaults to pt. A
% rule of thumb is that the baselineskip should be 1.2 times the font size.
\selectfont

\usepackage{parskip}
% Use the style of having no indentation with a space between paragraphs.

\usepackage{enumitem}
% Resume an enumerated list, continuing the old numbering, after some
% intervening text.

% \usepackage{fullpage}
% package that specifies normal margins

\usepackage{titling}
\setlength{\droptitle}{-5em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Macros

\newcommand{\chubby}[0]{\textsc{Chubby}}
\newcommand{\phat}[0]{\textsc{Phat}}
\newcommand{\phatraid}[0]{\textsc{PhatRaid}}
\newcommand{\raid}[1]{\textsc{RAID #1}}
\newcommand{\paxos}[0]{\textsc{Paxos}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title{PHAT RAID, yo}
\author{Andrew Johnson \and Daniel King \and Lucas Waye \and Scott Moore}
\date{May 13, 2014}

\maketitle

Distributed file stores, like Google's \chubby{} and CS260r's \phat{}, send all
files through a single master node which consistently replicates the file on
many slaves. In this scheme, the master node's throughput is a bottleneck for
storing large files. We present \phatraid{} which mitigates this bottleneck by
partitioning files across many \paxos{} groups which form a \raid{1} group.

\section{Introduction}

\section{Implementation}

\phatraid{} is implemented in Erlang, reusing our previous implementation of
\phat{}. The implementation is divided into three domains: the server cluster,
the client, and the raid-client. The server code implements a VR-like consensus
algorithm as well as simple file system. The client code exposes a file system
API which hides the communication bookkeeping. The raid-client provides a file
system API which transparently partitions files and reassembles them as they are
sent and received from the server cluster.

In Erlang, the basic unit of concurrency is a \emph{process} and an Erlang VM is
known as a \emph{node}. A process may communicate with other processes on its
node as well as processes on other nodes. Multiple nodes may be on the same
machine or distributed across many machines.

The server cluster is implemented as a collection of nodes running the same
server code. One of these nodes is designated the master node. The other nodes
are known as replicas.

The master and each replica has four components: the supervisor, the
(unfortunately named) server, VR, and the file system.

\begin{itemize}
\item the supervisor -- if any of the other three processes dies, it shuts all
  of them down and restarts all of them
\item the server -- if the node is the master, it passes messages on to VR; if
  the node is a replica, it redirects clients to the master
\item VR -- maintains a log of opaque messages, achieving consensus with other
  nodes in the cluster via the Viewstamped Replication protocol
\item the file system -- a simple hierarchical file system with locks
\end{itemize}

Each of these are implemented using Erlang behaviors. An Erlang behavior is a
framework which implements common patterns like servers and finite state
machines. In particular, the supervisor uses the \texttt{supervisor} behavior,
the server and file system both use the \texttt{gen\_server} behavior, and VR
uses the \texttt{gen\_fsm} behavior.

The \texttt{supervisor} behavior simply provides a framework for restarting failed
processes.

The \texttt{gen\_server} behavior provides a framework

\section{Performance}

\section{Conclusion}

\end{document}

### 2014-05-12 paper/paper.tex
Explain the three behaviors we use

@@
 \usepackage[sc,osf]{mathpazo}
-\linespread{1.05}         % Palatino needs more leading (space between lines)
+\linespread{1.05}         % Palatino needs more leading
 \usepackage[T1]{fontenc}
@@
 
-The \texttt{supervisor} behavior simply provides a framework for restarting failed
-processes.
+\begin{itemize}
+\item The \texttt{supervisor} behavior provides a framework for restarting
+  failed processes.
+\item The \texttt{gen\_server} behavior provides a framework for many-client
+  single-server interactions. We implemented functions to process messages
+  formatted as Erlang tuples. The \texttt{gen\_server} behavior handles
+  low-level socket interaction.
+\item The \texttt{gen\_fsm} behavior generalizes the \texttt{gen\_server}
+  behavior by permitting the server to have a finite number of states. Each
+  state has a set of functions with which to process messages. This behavior can
+  be seen as an ad hoc polymorphic variant of the \texttt{gen\_server} behavior.
+\end{itemize}
 
-The \texttt{gen\_server} behavior provides a framework
+Additionally, the \texttt{gen\_server} and \texttt{gen\_fsm} behaviors allow
+the programmer to specify an arbitrary state value which will be passed around
+\`{a} la functional reactive programming.
 

### 2014-05-12 paper/paper.tex
Add supervisor listing and discuss it

@@
 \usepackage{titling}
+% move the title up a bit, wastes less space
 \setlength{\droptitle}{-5em}
+
+\usepackage[usenames,dvipsnames]{color}
+% for \color used in listings
+\usepackage{textcomp}
+% for upquote used in listings
+\usepackage{listings}
+\lstset{
+    breakatwhitespace, % somehow prevents lstinline from line splitting
+    tabsize=2,
+    rulecolor=,
+    basicstyle=\footnotesize,
+    upquote=true,
+    aboveskip={1.5\baselineskip},
+    columns=fixed,
+    extendedchars=true,
+    breaklines=true,
+    prebreak = \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
+    frame=single,
+    showtabs=false,
+    showspaces=false,
+    showstringspaces=false,
+    keywordstyle=\color{Purple},
+    identifierstyle=\color{Black},
+    commentstyle=\color{BrickRed},
+    stringstyle=\color{RubineRed},
+    captionpos=b,
+}
+\lstloadlanguages{erlang}
+\lstset{language=erlang}
 
@@
 
+\lstinputlisting[firstline=6,lastline=18,float,
+                 caption=Supervisor Behavior --- \texttt{phat.erl},label=supervisor,
+                 numbers=left, firstnumber=6]
+                {../phat.erl}
+
+Listing \ref{supervisor} is taken from our supervisor code. When a supervisor is
+started, the \texttt{supervisor} behavior looks for a unary procedure called
+\lstinline!init!. The \lstinline!init! procedure returns a child specification which
+is a pair of a restart specification and a list of children. In Listing
+\ref{supervisor}, the restart specification, \lstinline!{one_for_all,1,5}!
+specifies that if any one node dies, all nodes should be restarted, unless more
+than one failure has occurred in the past five seconds. If more than one failure
+occurs in five seconds, the supervisor kills all the children and shuts itself
+down as well.
+
+Each child in the list of children is a 6-tuple consisting of: child name,
+initial procedure call, child transience, shutdown style, child type, necessary
+modules.
+
+\begin{itemize}
+\item The initial procedure call is a 3-tuple of module, procedure name and
+  argument list.
+\item Every child's transience is specified as \lstinline!permanent!, meaning
+  they should be restarted.
+\item Every child has shutdown style \lstinline!1! meaning they will first be
+  asked to terminate and one second later they will be forcibly terminated.
+\item All of our children are workers, which are distinguished from
+  sub-supervisors.
+\item The necessary modules are dynamically loaded when the child processes is
+  initialized, there is a one-to-one mapping between our children and code
+  modules.
+\end{itemize}
+
 \section{Performance}

### 2014-05-12 paper/paper.tex
More updates to paper

@@
 % package that allows you to change spacing
+
+\usepackage{cite}
+% for BibTeX
 
@@
 
+\usepackage{tikz}
+\usetikzlibrary{graphdrawing}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\subsection{System Organization}
+
 \phatraid{} is implemented in Erlang, reusing our previous implementation of
@@
 sent and received from the server cluster.
+
+\subsection{Erlang and Behaviors}
 
@@
 
-Each of these are implemented using Erlang behaviors. An Erlang behavior is a
-framework which implements common patterns like servers and finite state
+The four components are implemented using Erlang behaviors. An Erlang behavior
+is a framework which implements common patterns like servers and finite state
 machines. In particular, the supervisor uses the \texttt{supervisor} behavior,
@@
 
+\subsubsection{The \texttt{supervisor} Behavior}
+
 \lstinputlisting[firstline=6,lastline=18,float,
-                 caption=Supervisor Behavior --- \texttt{phat.erl},label=supervisor,
+                 caption=The \texttt{supervisor} Behavior --- \texttt{phat.erl},
+                 label=lst:supervisor,
                  numbers=left, firstnumber=6]
@@
 
-Listing \ref{supervisor} is taken from our supervisor code. When a supervisor is
-started, the \texttt{supervisor} behavior looks for a unary procedure called
-\lstinline!init!. The \lstinline!init! procedure returns a child specification which
-is a pair of a restart specification and a list of children. In Listing
-\ref{supervisor}, the restart specification, \lstinline!{one_for_all,1,5}!
-specifies that if any one node dies, all nodes should be restarted, unless more
-than one failure has occurred in the past five seconds. If more than one failure
-occurs in five seconds, the supervisor kills all the children and shuts itself
-down as well.
+Listing \ref{lst:supervisor} is taken from our supervisor code. When a
+supervisor is started, the \texttt{supervisor} behavior looks for a unary
+procedure called \lstinline!init!. The \lstinline!init! procedure returns a
+child specification which is a pair of a restart specification and a list of
+children. In Listing \ref{lst:supervisor}, the restart specification,
+\lstinline!{one_for_all,1,5}! specifies that if any one node dies, all nodes
+should be restarted, unless more than one failure has occurred in the past five
+seconds. If more than one failure occurs in five seconds, the supervisor kills
+all the children and shuts itself down as well.
 
@@
 
+\subsubsection{The \texttt{gen\_fsm} Behavior}
+
+\begin{lstlisting}[float,caption=The \texttt{gen\_fsm} Behavior --- \texttt{vr.erl},
+                   label=lst:genfsm, numbers=left, firstnumber=176]
+replica( {prepare, MasterViewNumber, _, _, _, _, _}
+       , State = #{ viewNumber := ViewNumber})
+    when MasterViewNumber > ViewNumber ->
+    ?debugFmt("my view is out of date, need to recover~n", []),
+    startRecovery(State);
+
+replica( {prepare, MasterViewNumber, _, _, _, _, _}
+       , State = #{ timeout := Timeout, viewNumber := ViewNumber})
+    when MasterViewNumber < ViewNumber ->
+    ?debugFmt("ignoring prepare from old view~n", []),
+    {next_state, replica, State, Timeout};
+
+replica( { prepare, _, Op, OpNumber
+         , MasterCommitNumber, Client, RequestNum}
+       , State = #{ prepareBuffer := PrepareBuffer }) ->
+    Message = {OpNumber, Op, Client, RequestNum},
+    NewPrepareBuffer = lists:sort([Message|PrepareBuffer]),
+    processPrepareOrCommit( OpNumber
+                          , MasterCommitNumber
+                          , NewPrepareBuffer
+                          , State
+                          );
+\end{lstlisting}
+
+\begin{lstlisting}[float,caption={Processing the VR Message Queue, Part 1 --- \texttt{vr.erl}},
+                   label=lst:processing1, numbers=left, firstnumber=375]
+processPrepareOrCommit( OpNumber, MasterCommitNumber, NewPrepareBuffer
+                      , State = #{ timeout := Timeout
+                                 , commitNumber := CommitNumber
+                                 , masterNode := MasterNode
+                                 , myNode := MyNode
+                                 , viewNumber := ViewNumber
+                                 , allNodes := Nodes})
+  when CommitNumber > MasterCommitNumber ->
+    NewViewNumber = ViewNumber + 1,
+    NewMaster = chooseMaster(State, NewViewNumber),
+    sendToReplicas( MasterNode
+                  , Nodes
+                  , {startViewChange, NewViewNumber, MyNode}
+                  ),
+    { next_state
+    , viewChange
+    , State#{ viewNumber := NewViewNumbern
+            , masterNode := NewMaster }
+    , Timeout
+    };
+\end{lstlisting}
+\begin{lstlisting}[float,caption={Processing the VR Message Queue, Part 2 --- \texttt{vr.erl}},
+                   label=lst:processing2, numbers=left, firstnumber=397]
+processPrepareOrCommit( OpNumber, MasterCommitNumber, NewPrepareBuffer
+                      , #{ timeout := Timeout } = State) ->
+    AfterBuffer =
+      processBuffer( State#{ prepareBuffer := NewPrepareBuffer }
+                   , NewPrepareBuffer
+                   , MasterCommitNumber
+                   ),
+    AfterLog = processLog(AfterBuffer, MasterCommitNumber),
+    #{ masterNode := MasterNode
+     , viewNumber := ViewNumber
+     , commitNumber := CommitNumber
+     , myNode := MyNode } = AfterLog,
+    if
+        CommitNumber < MasterCommitNumber ->
+            startRecovery(State);
+        true ->
+            sendToMaster( MasterNode
+                        , {prepareOk, ViewNumber, OpNumber, MyNode}
+                        ),
+            {next_state, replica, AfterLog, Timeout}
+    end.
+\end{lstlisting}
+
+A state in the \texttt{gen\_fsm} behavior manifests as a procedure which accepts
+two arguments: the incoming message and the store\footnote{Mutable state is
+  implemented as a pure store value which is passed around by the behavior}. The
+return value of a state-procedure is usually the 4-tuple
+\lstinline!{next_state,NextState,NewStore,Timeout}!. The \texttt{gen\_fsm}
+behavior will transition to \lstinline!NextState! and set the store to
+\lstinline!NewStore!. When a new message is received the \texttt{gen\_fsm}
+behavior will invoke the \lstinline!NextState! procedure with the new message
+and the \lstinline!NewStore!.
+
+We implemented Viewstamped Replication\cite{liskov2012viewstamped} to achieve
+consensus in the file system. Listing \ref{lst:genfsm} defines how to handle
+prepare messages sent to a node in the replica state. It is piecewise defined by
+pattern matching on the arguments. The first two definition clauses use the
+\lstinline!when! keyword to specify arbitrary, required relationships between
+the matched variables.
+
+The final definition clause triggers when the replica receives a message in the
+current view. The replica adds the message to a sorted queue and calls a
+processing procedure. The processing procedure is defined separately in Listing
+\ref{lst:processing1} and Listing \ref{lst:processing2}.
+
+Listing \ref{lst:processing1} handles a message from an out-of-date master,
+i.e., his commit number is older than the replica's commit number. The replica
+first proposes a view change to the cluster. Afterwards, it changes its
+\texttt{gen\_fsm} state to \lstinline!viewChange! and updates the
+store\footnote{The store is called \lstinline!State! in the listings} with the
+new view number and the new master.
+
+Listing \ref{lst:processing2} handles messages from the current master. The
+\lstinline!prepareBuffer! stores prepare messages that arrived out of order. In
+particular, if the log ends at operation number $n$ and message $n+1$ is dropped
+by the network, all subsequent messages will be buffered. The buffered messages
+will not be added to the log until message $n+1$ is received. The call to
+\lstinline!processBuffer! moves messages to the log, if all previous messages
+have now been received. The call to \lstinline!processLog! commits new messages
+in the log which have been committed by the master node. Lines 405 to 408
+destructure \lstinline!AfterLog! and bind variables for later use. The final
+\lstinline!if! statement responds to the master unless the master has committed
+messages we haven't received.\footnote{This can happen if the network drops a message whose
+  operation number lies between the replica's last commit number and the
+  master's last commit number.}
+
 \section{Performance}
@@
 
+\bibliographystyle{plain}
+\bibliography{paper}
+
 \end{document}

### 2014-05-13 paper/paper.tex
Add a system organization section

@@
 \usepackage{tikz}
-\usetikzlibrary{graphdrawing}
+\usetikzlibrary{backgrounds}
+\usetikzlibrary{fit}
+\usetikzlibrary{positioning}
+
+\pgfdeclarelayer{b}
+\pgfdeclarelayer{f}
+\pgfsetlayers{b,main,f}
+
+\newcommand{\plate}[4]{
+\begin{pgfonlayer}{b}
+\node (invis#1) [draw, color=white, inner sep=1pt,rectangle,fit=#2] {};
+\end{pgfonlayer}
+\begin{pgfonlayer}{f}
+\node (capt#1) [ below left=0 pt of invis#1.south east, xshift=2pt,yshift=1pt, fill=black!15]
+{\footnotesize{#3}};
+\end{pgfonlayer}
+\begin{pgfonlayer}{b}
+\node (#1) [fill=black!15, rounded corners, inner sep=5pt, rectangle,fit=(invis#1) (capt#1),#4] {};
+\end{pgfonlayer}
+}
 
@@
 \newcommand{\paxos}[0]{\textsc{Paxos}}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{figure}
+  \begin{tikzpicture}
+    [->,scale=.8,]
+    \node [rectangle, draw, thick] (raidclient) at (0,2) {Raid Client};
+    \node [rectangle, draw, thick] (client1) at (-4,1) {Client 1};
+    \node [rectangle, draw, thick] (client2) at (0,0) {Client 2};
+    \node [rectangle, draw, thick] (client3) at (4,1) {Client 3};
+    \node [rectangle, draw, thick] (vr11) at (-7, -2) {Master};
+    \node [rectangle, draw, thick] (vr12) at (-5, -2) {Replica};
+    \node [rectangle, draw, thick] (vr13) at (-3, -2) {Replica};
+    \node [rectangle, draw, thick] (vr21) at (-2, -4) {Master};
+    \node [rectangle, draw, thick] (vr22) at (0, -4) {Replica};
+    \node [rectangle, draw, thick] (vr23) at (2, -4) {Replica};
+    \node [rectangle, draw, thick] (vr31) at (3, -2) {Master};
+    \node [rectangle, draw, thick] (vr32) at (5, -2) {Replica};
+    \node [rectangle, draw, thick] (vr33) at (7, -2) {Replica};
+
+    \foreach \to in {client1, client2, client3}
+    \draw [thick] (raidclient) -- (\to);
+    \foreach \to in {vr11, vr12, vr13}
+    \draw [thick] (client1) -- (\to);
+    \foreach \to in {vr21, vr22, vr23}
+    \draw [thick] (client2) -- (\to);
+    \foreach \to in {vr31, vr32, vr33}
+    \draw [thick] (client3) -- (\to);
+
+    \plate{clientnode}{(raidclient)(client1)(client2)(client3)}{Client Node}{}
+    \plate{pc1}{(vr11)(vr12)(vr13)}{Phat Cluster \#1}{}
+    \plate{pc2}{(vr21)(vr22)(vr23)}{Phat Cluster \#2}{}
+    \plate{pc3}{(vr31)(vr32)(vr33)}{Phat Cluster \#3}{}
+
+  \end{tikzpicture}
+  \caption{Organizational Diagram}
+  \label{fig:organization}
+\end{figure}
+
 \phatraid{} is implemented in Erlang, reusing our previous implementation of
@@
 sent and received from the server cluster.
+
+Figure \ref{fig:organization} depicts the organization of our system. The
+raid-client delegates communication to each client. Each client is connected to
+one \phat{} cluster, which has a master and a number of replicas.
+
+Each \phat{} cluster maintains a file system which is separate and distinct from
+the file systems of other \phat{} clusters. Generally, two phat clusters do not
+communicate.
+
+\subsection{Implenting RAID on \phat{}}
 
@@
 pattern matching on the arguments. The first two definition clauses use the
-\lstinline!when! keyword to specify arbitrary, required relationships between
-the matched variables.
+\lstinline!when! keyword to specify arbitrary required relationships between the
+matched variables.
 

### 2014-05-13 paper/paper.tex
some stuff on hdd vs sdd

@@
 \usepackage{amsmath}
+\usepackage{url}
 \usepackage{amssymb}
@@
 \section{Introduction}
+
+\subsection{Disk Performance}
+
+An important consideration in our project is the impact disk performance has on overall commit latency.
+We believe that disk read/write latency is the major contributor to storing large files in a distributed system.
+{\bf TODO more}
+
+In a review of hard disk performance \cite{disk-perf}, Hard Disk Drives (HDD) were compared with Solid State Drives (SDD).
+HDDs use a physical magnetic platter that spins so that a drive head can read or write data by sensing or manipulating the magnetic field in a particular position on to the platter.
+In contrast, Solid SDD have no moving parts and instead uses integrated circuits as memory to store data persistently.
+Because the way in which data is accessed is different, the performance characteristics between the two are different.
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{disk-perf}
+\caption{{\bf Left Chart}: Solid State Disk reads across entire drive (Left -- Outermost sector, Right -- Innermost sector), {\bf Right Chart}: Hard Disk reads across the entire drive (Left -- Outer rim, Right -- Inner rim).
+From \cite{disk-perf}}
+\label{fig:hdd-vs-sdd}
+\end{figure}
+
+Figure \ref{fig:hdd-vs-sdd} shows a comparison between read times for a SDD and HDD as a function of position on the disk.
+HDD storage performance is greatly influenced by where on the disk platter the data was placed.
+Data can be placed on the outer rim of the platter much more quickly than data placed near the center of the platter as the outer rim is moving faster in comparison (due to conservation of angular momentum).
+In contrast to HDDs, SDDs do not exhibit this problem as there is no spinning platter.
+The right diagram in \ref{fig:hdd-vs-sdd} shows relatively constant read performance throughout the entire disk.
+
+\urldef\hwbenchmark\url{http://www.tomshardware.com/charts/ssd-charts-2013/benchmarks%2C129.html}
+\begin{figure}
+\centering
+\begin{tabular}{l || c}
+\multicolumn{2}{c}{{\bf Solid State Disk Performance Benchmarks}} \\ \hline
+Sequential Write Speed  \hspace{7em} & 55.03 - 481.85 MB/sec \\ 
+Write Access Time & 0.03 - 0.32 seconds \\ \hline
+Sequential Read Speed &  207.95 - 522.45 MB/sec \\
+Read Access Time & 0.03 - 0.22 seconds
+\end{tabular}
+\caption[Solid State Performance]{Solid State Disk performance for middle 95\% of sold state drives tested in a benchmark. Benchmark available at Benchmark available at \hwbenchmark.}
+\label{fig:sdd-perf}
+\end{figure}
+
+Figure \ref{fig:sdd-perf} shows the performance for SSDs from a 2013 survey of commercially available SDDs.
 

### 2014-05-13 paper/paper.tex
Incomplete updates to evaluation

@@
 
+\usepackage{caption}
+\usepackage{subcaption}
+% for subfigures
+
 \usepackage[usenames,dvipsnames]{color}
@@
 \usetikzlibrary{positioning}
+\usetikzlibrary{shapes.misc} % for cross shaped nodes
 
@@
 \newcommand{\phatraid}[0]{\textsc{PhatRaid}}
+\newcommand{\phatraidcf}[2]{\textsc{PhatRaid}(#1,#2)}
 \newcommand{\raid}[1]{\textsc{RAID #1}}
@@
 
+\subsection{Erlang Primer}
+
+\phatraid{} is implemented in Erlang. Erlang is a pure, functional language with
+extensive library support for distributed applications. Erlang has one unit of
+concurrency, the process, and one unit of distribution, the node. Each Erlang
+node consists of an arbitrary number of processes. Erlang nodes may live on the
+same machine or on different machines connected by a network. Every Erlang node
+is given a name by an Erlang name server. Erlang processes may communicate with
+any Erlang node which they can name. Erlang processes interact with one another
+by making RPC calls.
+
 \subsection{System Organization}
@@
 \begin{figure}
+  \centering
   \begin{tikzpicture}
-    [->,scale=.8,]
-    \node [rectangle, draw, thick] (raidclient) at (0,2) {Raid Client};
-    \node [rectangle, draw, thick] (client1) at (-4,1) {Client 1};
-    \node [rectangle, draw, thick] (client2) at (0,0) {Client 2};
-    \node [rectangle, draw, thick] (client3) at (4,1) {Client 3};
-    \node [rectangle, draw, thick] (vr11) at (-7, -2) {Master};
-    \node [rectangle, draw, thick] (vr12) at (-5, -2) {Replica};
-    \node [rectangle, draw, thick] (vr13) at (-3, -2) {Replica};
-    \node [rectangle, draw, thick] (vr21) at (-2, -4) {Master};
-    \node [rectangle, draw, thick] (vr22) at (0, -4) {Replica};
-    \node [rectangle, draw, thick] (vr23) at (2, -4) {Replica};
-    \node [rectangle, draw, thick] (vr31) at (3, -2) {Master};
-    \node [rectangle, draw, thick] (vr32) at (5, -2) {Replica};
-    \node [rectangle, draw, thick] (vr33) at (7, -2) {Replica};
+    [<->,scale=.85,shorten <=2pt, shorten >=2pt,]
+    \node [rectangle, draw, thick, inner sep=5pt] (raidclient) at (0,2) {Raid Client};
+    \node [rectangle, draw, thick, inner sep=5pt] (client1) at (-4,1) {Client 1};
+    \node [rectangle, draw, thick, inner sep=5pt] (client2) at (0,0) {Client 2};
+    \node [rectangle, draw, thick, inner sep=5pt] (client3) at (4,1) {Client 3};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr11) at (-6, -2) {Replica};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr12) at (-4, -2) {Master};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr13) at (-2, -2) {Replica};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr21) at (-2, -4) {Replica};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr22) at (0, -4) {Master};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr23) at (2, -4) {Replica};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr31) at (2, -2) {Replica};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr32) at (4, -2) {Master};
+    \node [rectangle, draw, thick, inner sep=5pt] (vr33) at (6, -2) {Replica};
 
@@
       \draw [thick] (raidclient) -- (\to);
-    \foreach \to in {vr11, vr12, vr13}
-    \draw [thick] (client1) -- (\to);
-    \foreach \to in {vr21, vr22, vr23}
-    \draw [thick] (client2) -- (\to);
-    \foreach \to in {vr31, vr32, vr33}
-    \draw [thick] (client3) -- (\to);
+    \draw [thick] (client1) -- (vr12);
+    \draw [thick] (client2) -- (vr22);
+    \draw [thick] (client3) -- (vr32);
+
+    % \foreach \to in {vr11, vr13}
+    %   \draw [thick] (vr12) -- (\to);
+    % \foreach \to in {vr21, vr23}
+    %   \draw [thick] (vr22) -- (\to);
+    % \foreach \to in {vr31, vr33}
+    %   \draw [thick] (vr32) -- (\to);
 
@@
   \end{tikzpicture}
-  \caption{Organizational Diagram}
+  \caption{A \phatraidcf{3}{1} Group}
   \label{fig:organization}
@@
 
-\subsection{Implenting RAID on \phat{}}
+\subsection{RAIDing \phat{}}
 
-\subsection{Erlang and Behaviors}
+A \phatraidcf{$C$}{$f$} implementation is parameterized by two numbers: $C$, the
+number of \phat{} clusters, and $f$, the \phat{} cluster failure threshold. A
+single \phat{} cluster contains $2f + 1$ nodes. A \phatraid{} group contains $C$
+clusters for a total of $C(2f+1)$ \phat{} nodes. A \phatraid{} group can be
+concurrently accessed by an arbitrary number of raid-clients.
+
+On a client node, the raid-client spawns $C$ client processes which each maintain
+a connection to one of the $C$ \phat{} clusters. All communications between the
+raid-client and a phat cluster go through the assigned client process.
+
+\subsection{Erlang Behaviors}
 
@@
 \texttt{gen\_fsm} state to \lstinline!viewChange! and updates the
-store\footnote{The store is called \lstinline!State! in the listings} with the
-new view number and the new master.
+store\footnote{The store is called \lstinline!State! in the code listings} with
+the new view number and the new master.
 
@@
 
-\section{Performance}
+\section{Evaluation}
+
+\subsection{Resilience}
+
+We distinguish between two types of failure: cessation of progress and loss of
+data. In a VR cluster without persistent stores, data loss only occurs when all
+$2f+1$ nodes in the cluster fail. In a VR cluster, progress only ceases if $f+1$
+nodes are lost. In \phatraid{} the conditions for cessation of progress and loss
+of data are complicated by the combination of RAID and VR.
+
+\begin{figure}
+  \centering
+  \begin{subfigure}{0.4\textwidth}
+    \centering
+    \begin{tikzpicture}[<->,scale=.5,shorten <=2pt, shorten >=2pt,]
+      \foreach \x/\y in {1/1, 1/2, 1/3, 2/1, 2/2, 2/3, 3/1, 3/2, 3/3}
+        \node [circle, draw, thick] (pr\x\y) at (\x,-\y) {};
+      \foreach \x/\y in {3/1, 3/2, 3/3, 1/3, 2/3}
+        \node [strike out, draw=red, thick] (pr\x\y) at (\x,-\y) {};
+    \end{tikzpicture}
+    \caption{Progress, No Data Loss}
+    \label{fig:failures_allok}
+  \end{subfigure}
+~
+  \begin{subfigure}{0.4\textwidth}
+    \centering
+    \begin{tikzpicture}[<->,scale=.5,shorten <=2pt, shorten >=2pt,]
+      \foreach \x/\y in {1/1, 1/2, 1/3, 2/1, 2/2, 2/3, 3/1, 3/2, 3/3}
+        \node [circle, draw, thick] (pr\x\y) at (\x,-\y) {};
+      \foreach \x/\y in {1/1, 1/2, 2/1, 2/2}
+        \node [strike out, draw=red, thick] (pr\x\y) at (\x,-\y) {};
+    \end{tikzpicture}
+    \caption{No Progress, No Data Loss}
+    \label{fig:failures_noprogress}
+  \end{subfigure}
+\\\vspace{1em}
+  \begin{subfigure}{0.4\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=.5,shorten <=2pt, shorten >=2pt,]
+      \node[rotate=25] (N) at (1.5,-1.5) {\textsc{Impossible!}};
+    \end{tikzpicture}
+    \caption{Progress, Data Loss}
+    \label{fig:failures_max}
+  \end{subfigure}
+~
+  \begin{subfigure}{0.4\textwidth}
+    \centering
+    \begin{tikzpicture}[<->,scale=.5,shorten <=2pt, shorten >=2pt,]
+      \foreach \x/\y in {1/1, 1/2, 1/3, 2/1, 2/2, 2/3, 3/1, 3/2, 3/3}
+        \node [circle, draw, thick] (pr\x\y) at (\x,-\y) {};
+      \foreach \x/\y in {1/1, 1/2, 1/3, 2/1, 2/2, 2/3}
+        \node [strike out, draw=red, thick] (pr\x\y) at (\x,-\y) {};
+    \end{tikzpicture}
+    \caption{No Progress, Data Loss}
+    \label{fig:failures_noprogress_nodata}
+  \end{subfigure}
+
+  \caption{Some Failed Node Distributions of \phatraidcf{3}{1}}
+  \label{fig:failures}
+\end{figure}
+
+Four distributions of failed nodes in \phatraidcf{3}{1} are depicted in Figure
+\ref{fig:failures}. In this section, if we claim ``data is lost'', we mean that
+if no persistent store is used, data will be lost. Adding a persistent store
+obviously prevents loss of committed data.
+
+The top left corner depicts five of the nine nodes failing. This state is
+\emph{not} a failure state. The group can still make progress and no data is
+lost. The two surviving clusters can reach consensus on new values and can
+reconstruct the failed cluster's data.
+
+The top right corner depicts four of the nine nodes failing. This state is a
+failure state. The group cannot make progress but no data can been
+lost. Progress is blocked because more than one cluster cannot reach
+consensus. Since no cluster was completely destroyed, all data still exists on
+at least one node.
+
+The bottom left corner is impossible because if data can be lost then we
+certainly cannot make progress.
+
+The bottom right corner depicts six of the nine nodes failing. This state is
+also a failure state. The group cannot make progress and will lose
+data. Progress is blocked because more than one cluster cannot reach
+consensus. Additionally, two complete clusters have been lost so half of the
+data cannot be recovered.
+
+In general, progress ceases if more than one cluster cannot reach consensus and
+data loss occurs if more than one cluster completely fails. The lower bound of
+failed nodes for progress cessation in \phatraidcf{$C$}{$f$} is:
+    $$ C\cdot(f+1) $$
+and the lower bound of failed nodes for data loss is:
+    $$ 2\cdot(2f+1) $$
+
+In contrats, if all data has been propagated to all nodes, data is
+only lost when all nodes die. If data has not been propagated to all nodes, then
+loss occurs if all nodes containing the data die.
+
+\subsection{Performance}
 

### 2014-05-13 paper/paper.tex
hdd perf

@@
 
-In a review of hard disk performance \cite{disk-perf}, Hard Disk Drives (HDD) were compared with Solid State Drives (SDD).
+In a review of hard disk performance \cite{disk-perf}, Hard Disk Drives (HDD) were compared with Solid State Drives (SSD).
 HDDs use a physical magnetic platter that spins so that a drive head can read or write data by sensing or manipulating the magnetic field in a particular position on to the platter.
-In contrast, Solid SDD have no moving parts and instead uses integrated circuits as memory to store data persistently.
+In contrast, SSDs have no moving parts and instead uses integrated circuits as memory to store data persistently.
 Because the way in which data is accessed is different, the performance characteristics between the two are different.
@@
 
-Figure \ref{fig:hdd-vs-sdd} shows a comparison between read times for a SDD and HDD as a function of position on the disk.
+Figure \ref{fig:hdd-vs-sdd} shows a comparison between read times for a SSD and HDD as a function of position on the disk.
 HDD storage performance is greatly influenced by where on the disk platter the data was placed.
 Data can be placed on the outer rim of the platter much more quickly than data placed near the center of the platter as the outer rim is moving faster in comparison (due to conservation of angular momentum).
-In contrast to HDDs, SDDs do not exhibit this problem as there is no spinning platter.
+In contrast to HDDs, SSDs do not exhibit this problem as there is no spinning platter.
 The right diagram in \ref{fig:hdd-vs-sdd} shows relatively constant read performance throughout the entire disk.
+Additionally, HDDs have difficultly with random accesses as the drive may need to wait for the drive head to be directly over the correct position on the platter.
+This problem becomes especially apparent when an individual file is fragmented into multiple blocks by the file system.
+SSDs have relatively better random access times, in general, though sequential access tends to be many times faster than random access \cite{disk-perf}.
+HDD performance can also can suffer from {\em spin-up delay}, which happens as a result of a power-saving measure that stops spinning the disk during idle times.
+
+We decided that HDDs have far too much performance variability in order to be properly modeled.
+As a result, we instead chose to simulate the performance characteristics of SSDs as they have relatively constant access times regardless of sector position on the drive and also do not suffer from {\em spin-up delay}.
 
@@
 Sequential Write Speed  \hspace{7em} & 55.03 - 481.85 MB/sec \\ 
-Write Access Time & 0.03 - 0.32 seconds \\ \hline
+Write Access Time & 0.03 - 0.32 ms \\ \hline
 Sequential Read Speed &  207.95 - 522.45 MB/sec \\
-Read Access Time & 0.03 - 0.22 seconds
+Read Access Time & 0.03 - 0.22 ms
 \end{tabular}
@@
 
-Figure \ref{fig:sdd-perf} shows the performance for SSDs from a 2013 survey of commercially available SDDs.
+Figure \ref{fig:sdd-perf} shows the performance for SSDs from a 2013 survey of commercially available SSDs.
+Sequential write speeds measure the throughput (in megabytes per second) of data stored on the drive when the data is being written in the drive's natural sequential order (e.g.
+increasing sectors).
+Similarly, sequential read speed measures the throughput for sequential reads.
+Read/Write access times represent the latency of performing a single random read or write.
 

### 2014-05-13 paper/paper.tex
graphics

@@
 This problem becomes especially apparent when an individual file is fragmented into multiple blocks by the file system.
-SSDs have relatively better random access times, in general, though sequential access tends to be many times faster than random access \cite{disk-perf}.
+SSDs have relatively better random access times, in general, than HDDs.
+Sequential access tends to be many times faster than random access for both SSDs and HDDs \cite{disk-perf}.
 HDD performance can also can suffer from {\em spin-up delay}, which happens as a result of a power-saving measure that stops spinning the disk during idle times.
@@
 Sequential Read Speed &  207.95 - 522.45 MB/sec \\
-Read Access Time & 0.03 - 0.22 ms
+Read Access Time & 0.03 - 0.22 ms \\ \hline
 \end{tabular}
-\caption[Solid State Performance]{Solid State Disk performance for middle 95\% of sold state drives tested in a benchmark. Benchmark available at Benchmark available at \hwbenchmark.}
+\caption[Solid State Performance]{Solid State Disk performance for middle 95\% of solid state drives tested in a benchmark. Benchmark available at Benchmark available at \hwbenchmark.}
 \label{fig:sdd-perf}
@@
 communicate.
+
+\subsection{Disk Emulation and Abstraction Layer}
+
+We developed a novel ``{\bf {\em Disk Emulation and Abstraction Layer}}'' (DEAL) to simulate SSD performance so they .
 

### 2014-05-13 clean-and-launch-vr.sh
strangeness

@@ -1,0 +1,24 @@
+N=$1
+
+[ "$#" -eq 1 ] || die "1 arguments required, $# provided. Valid invocation:
+
+  bash clean-and-launch-raid.sh N C
+
+  - N -- the total number of nodes
+"
+
+rm -r tests
+mkdir tests
+killall beam.smp
+erlc *.erl
+bash start-raid.sh $N 1 tests
+NODE_LIST+="["
+for j in $(seq 1 $(expr $N - 1))
+do
+    NODE_LIST+="n$j@localhost, "
+done
+NODE_LIST+="n$N@localhost]"
+
+echo erl -sname foo@localhost -eval "client:start_link($NODE_LIST)"
+erl -sname foo@localhost -eval "client:start_link($NODE_LIST)"
+

### 2014-05-14 clean-and-launch-raid.sh
Fix launch script

@@ -14,18 +14,19 @@ mkdir tests
 killall beam.smp
 erlc *.erl
 bash start-raid.sh $N $C tests
+NODE_PER_CLUSTER=$(expr $N / $C)
 NODE_LIST="["
-for i in $(seq 1 $C $(expr $N - $C))
+for i in $(seq 1 $NODE_PER_CLUSTER $(expr $N - $NODE_PER_CLUSTER))
 do
     NODE_LIST+="["
-    for j in $(seq $i $(expr $i + $C - 2))
+    for j in $(seq $i $(expr $i + $NODE_PER_CLUSTER - 2))
     do
         NODE_LIST+="n$j@localhost, "
     done
     NODE_LIST+="n$(expr $j + 1)@localhost],"
 done
 NODE_LIST+="["
-for j in $(seq $(expr $N - $C + 1) $(expr $N - 1))
+for j in $(seq $(expr $N - $NODE_PER_CLUSTER + 1) $(expr $N - 1))
 do
     NODE_LIST+="n$j@localhost, "
 done

### 2014-05-14 paper/paper.tex
spelling fix
line was too long
Add a theoretical analysis of failure probabilities

@@
 
-We developed a novel ``{\bf {\em Disk Emulation and Abstraction Layer}}'' (DEAL) to simulate SSD performance so they .
+We developed a novel ``{\bf {\em Disk Emulation and Abstraction Layer}}'' (DEAL)
+to simulate SSD performance so they .
 
@@
     \caption{Progress, Data Loss}
-    \label{fig:failures_max}
+    \label{fig:failures_impossible}
   \end{subfigure}
@@
     \caption{No Progress, Data Loss}
-    \label{fig:failures_noprogress_nodata}
+    \label{fig:failures_noprogress_dataloss}
   \end{subfigure}
@@
 failed nodes for progress cessation in \phatraidcf{$C$}{$f$} is:
-    $$ C\cdot(f+1) $$
+    $$ 2\cdot(f+1) $$
 and the lower bound of failed nodes for data loss is:
@@
 
-In contrats, if all data has been propagated to all nodes, data is
-only lost when all nodes die. If data has not been propagated to all nodes, then
-loss occurs if all nodes containing the data die.
+In contrast, if all data has been propagated to all nodes, data is only lost
+when all nodes die. If data has not been propagated to all nodes, then loss
+occurs if all nodes containing the data die.
+
+\subsubsection{Theoretical Analysis}
+
+A single number cannot describe the failure threshold of a \phatraid{} cluster
+because of the interplay between VR and RAID. The failure of the \phatraid{}
+group is dependent on the arrangement of the failed nodes. For example, Figure
+\ref{fig:failures_noprogress} depicts a situation wherein only four failures
+stops progress. In contrast, Figure \ref{fig:failures_allok} depicts a situation
+wherein five failures does \emph{not} stop progress.
+
+In a \phatraidcf{3,1} group, less than one-quarter of four-failure arrangements
+stop progress. The necessary condition for progress cessation is that two or
+more clusters have lost at least $f+1$ nodes. By the pidgeon hole principle,
+only the two-in-two, four-failure arrangement stops progress. We can count the
+number of such arrangements:
+
+    $$ \binom{3}{2} \binom{3}{2}^2 = 3^3 = 27 $$
+
+The first term represents choosing the two clusters of three that will
+fail. The second term represents choosing which two nodes in each cluster of
+three nodes will fail.
+
+The total number of four-failure arrangements is simply:
+
+    $$ \binom{9}{4} = 126 $$
+
+If all nodes fail uniformly and independently, then the probability of four
+nodes causing \phatraid{} group failure is
+
+    $$ \frac{27}{126} \approx 0.21 $$
+
+When five nodes fail, they can be distributed into each group in the
+proportions: (3,1,1), (3,2,0) and (2,2,1). The first case does not stop progress
+because only one cluster fails. The later two cases stop progress because two
+clusters fail. We calculate the probability of \phatraid{} group failure when
+five nodes fail:
+
+\begin{align*}
+  \textrm{(3,1,1)}\;\;\;
+    &\binom{3}{1}\binom{3}{3}\binom{3}{1}\binom{3}{1} = 3^3 = 27 \\
+  \textrm{(3,2,0)}\;\;\;
+    &\binom{3}{1}\binom{2}{1}\binom{3}{3}\binom{3}{2} = 2\cdot3^2 = 18 \\
+  \textrm{(2,2,1)}\;\;\;
+    &\binom{3}{1}\binom{3}{2}\binom{3}{1}\binom{3}{1} = 3^4 = 81 \\
+  \textrm{5-Failure Arrs.}\;\;\;
+    &\binom{9}{5} = 126 \\
+  \textrm{Failure Prob.}\;\;\;
+    &\Pr\left[\textrm{group failure} \,|\, n_{failed} = 5\right] = \frac{81 + 18}{126} =
+    \frac{99}{126} \approx 0.79 \\
+\end{align*}
+
+A VR or PAXOS cluster of size nine is resilient to $f = 4$ failures. In
+contrast, \phatraidcf{3}{1} is absolutely resilient to $f=3$ failures, has high
+probability of resilience to $f=4$ failures, and has low probability of
+resilience to $f=5$ failures.
+
+The general formula for the failure probability distribution is not clear, but
+it is clear that a \phatraidcf{$C$}{$f$} group is absolutely resilient to $2(f +
+1) - 1$ or fewer failures. A \phatraidcf{$C$}{$f$} group is likely resilient
+to $2(f+1)$ failures and unlikely resistant to $2(f+1)+1$ failures.
+
+The general form for the probability of \phatraidcf{$C$}{$f$} group failure
+given $2(f + 1)$ node failures is:
+
+\begin{align*}
+  \Pr\left[\textrm{group failure} \,|\, \phatraidcf{C}{f}, n_{failure}=2(f+1)\right]
+    =
+  \frac{\binom{C}{2}\binom{2f+1}{f+1}^2}
+       {\binom{C \cdot (2f+1)}{2(f+1)}}
+\end{align*}
+
+For \phatraidcf{6}{1} we calculate:
+
+\begin{align*}
+  P\left[\textrm{group failure} \,|\, \phatraidcf{6}{1}, n_{failure}=4\right]
+    = \frac{135}{3000} \approx 0.05
+\end{align*}
 
 \subsection{Performance}
+
+\section{Future Work}
+
+Extra parity bits!
 

### 2014-05-14 paper/plots.R
Performance graphs

@@ -1,0 +1,12 @@
+library(ggplot2)
+times <- read.csv("~/Documents/schoolwork/cs260/erlang-phat/paper/times.csv")
+
+ggplot(times, aes(N..total.nodes., Store.Time..ms., label = Label, group = Group)) + 
+  geom_line() +
+  geom_text(vjust=-1.25,size=4)+ ylim(0,2000) + xlim(0, 35) + 
+  theme_bw() + xlab("Number of Total Nodes") + ylab("Store Time (ms)") + geom_point()
+
+ggplot(times, aes(N..total.nodes., Fetch.Time..ms., label = Label, group = Group)) + 
+  geom_line() +
+  geom_text(vjust=-1.25,size=4)+ ylim(0,2000) + xlim(0, 35) + 
+  theme_bw() + xlab("Number of Total Nodes") + ylab("Fetch Time (ms)") + geom_point()

### 2014-05-14 paper/paper.tex
Bunch of changes to the paper

@@
 
+Google's \chubby{}\cite{burrows-chubby} is a distributed file system which
+achieves consensus in the face of limited failures. Unfortunately, for storing
+large files disk latency on the cluster nodes is a bottleneck. We propose
+building a RAID group out of distributed file system clusters. Large files are
+sharded and disk latency is reduced proportional to number of clusters. Failure
+resiliency is not completely sacrificed and we provide an analysis of the
+new failure modes.
+
 \subsection{Disk Performance}
@@
 extensive library support for distributed applications. Erlang has one unit of
-concurrency, the process, and one unit of distribution, the node. Each Erlang
-node consists of an arbitrary number of processes. Erlang nodes may live on the
-same machine or on different machines connected by a network. Every Erlang node
-is given a name by an Erlang name server. Erlang processes may communicate with
-any Erlang node which they can name. Erlang processes interact with one another
-by making RPC calls.
+concurrency, the \emph{process}, and one unit of distribution, the
+\emph{node}. Each Erlang node consists of an arbitrary number of
+processes. Erlang nodes may live on the same machine or on different machines
+connected by a network. Every Erlang node is given a name by an Erlang name
+server. Erlang processes may communicate with any Erlang node or process it can
+name. Erlang processes interact with one another by making RPC calls or sending
+messages. Identifiers beginning with a lowercase letter are actually atoms, like
+Scheme's symbols. Identifiers beginning with an uppercase letter are normal
+identifiers.
 
@@
 the file systems of other \phat{} clusters. Generally, two phat clusters do not
-communicate.
+communicate. The raid-client and client processes all live on one node. Each
+\phat{} master and \phat{} replica lives on its own node.
 
@@
 A \phatraidcf{$C$}{$f$} implementation is parameterized by two numbers: $C$, the
-number of \phat{} clusters, and $f$, the \phat{} cluster failure threshold. A
+number of \phat{} clusters, and $f$, the \phat{} cluster failure tolerance. A
 single \phat{} cluster contains $2f + 1$ nodes. A \phatraid{} group contains $C$
@@
 
-On a client node, the raid-client spawns $C$ client processes which each maintain
-a connection to one of the $C$ \phat{} clusters. All communications between the
-raid-client and a phat cluster go through the assigned client process.
+On a client node, the raid-client spawns $C$ client processes which each
+maintain a connection to one of the $C$ \phat{} clusters. Each client process
+mediates all communication between its assigned \phat{} cluster and the
+raid-client.
+
+A store operation on the raid-client breaks the file into $C-1$ chunks, stores
+each chunk on a cluster, and stores a parity bit on the remaining cluster. A
+raid-client fetch operation reconstructs the file from the $C-1$ chunks, using
+the parity bit if necessary.
 
@@
 
-In Erlang, the basic unit of concurrency is a \emph{process} and an Erlang VM is
-known as a \emph{node}. A process may communicate with other processes on its
-node as well as processes on other nodes. Multiple nodes may be on the same
-machine or distributed across many machines.
-
-The server cluster is implemented as a collection of nodes running the same
-server code. One of these nodes is designated the master node. The other nodes
-are known as replicas.
+The server cluster is a collection of nodes running the same server code. One of
+these nodes is designated the master node (aka \phat{} Master). The other nodes are
+known as \phat{} replicas or just replicas.
 
 The master and each replica has four components: the supervisor, the
-(unfortunately named) server, VR, and the file system.
+(unfortunately named) server, VR, and the file system:
 
 \begin{itemize}
-\item the supervisor -- if any of the other three processes dies, it shuts all
-  of them down and restarts all of them
-\item the server -- if the node is the master, it passes messages on to VR; if
-  the node is a replica, it redirects clients to the master
+\item the supervisor -- restarts the other three processes if any single node dies
+\item the server -- passes messages to the VR process if this node is the
+  \phat{} master; otherwise, it drops the message and informs the client of the
+  current master node
 \item VR -- maintains a log of opaque messages, achieving consensus with other
   nodes in the cluster via the Viewstamped Replication protocol
-\item the file system -- a simple hierarchical file system with locks
+\item the file system -- maintains a simple hierarchical file system with locks
 \end{itemize}
@@
   formatted as Erlang tuples. The \texttt{gen\_server} behavior handles
-  low-level socket interaction.
+  low-level socket interaction, messaging queueing, etc.
 \item The \texttt{gen\_fsm} behavior generalizes the \texttt{gen\_server}
@@
 Each child in the list of children is a 6-tuple consisting of: child name,
-initial procedure call, child transience, shutdown style, child type, necessary
-modules.
+initial procedure call, child transience, shutdown style, child type, and
+necessary modules.
 
@@
 \item Every child's transience is specified as \lstinline!permanent!, meaning
-  they should be restarted.
+  they \emph{should} be restarted.
 \item Every child has shutdown style \lstinline!1! meaning they will first be
-  asked to terminate and one second later they will be forcibly terminated.
+  asked to terminate and one second later they will be forcibly
+  terminated\footnote{We could also have specified the hastier shutdown style:
+    \lstinline!brutal_kill!}.
 \item All of our children are workers, which are distinguished from
@@
 \item The necessary modules are dynamically loaded when the child processes is
-  initialized, there is a one-to-one mapping between our children and code
-  modules.
+  initialized. In \phatraid{}, each child only depends on one module.
 \end{itemize}
@@
 
-We implemented Viewstamped Replication\cite{liskov2012viewstamped} to achieve
-consensus in the file system. Listing \ref{lst:genfsm} defines how to handle
-prepare messages sent to a node in the replica state. It is piecewise defined by
-pattern matching on the arguments. The first two definition clauses use the
-\lstinline!when! keyword to specify arbitrary required relationships between the
-matched variables.
+The \phat{} file system achieves consensus via the Viewstamped Replication
+protocol \cite{liskov2012viewstamped}. Listing \ref{lst:genfsm} defines how to
+handle prepare messages sent to a node in the replica state. It is piecewise
+defined by pattern matching on the arguments. The first two definition clauses
+use the \lstinline!when! keyword to specify arbitrary required relationships
+between the matched variables.
 
@@
 current view. The replica adds the message to a sorted queue and calls a
-processing procedure. The processing procedure is defined separately in Listing
-\ref{lst:processing1} and Listing \ref{lst:processing2}.
+processing procedure. The processing procedure is defined separately in Listings
+\ref{lst:processing1} and \ref{lst:processing2}.
 
 Listing \ref{lst:processing1} handles a message from an out-of-date master,
-i.e., his commit number is older than the replica's commit number. The replica
-first proposes a view change to the cluster. Afterwards, it changes its
+i.e., the master's commit number is older than the replica's commit number. The
+replica first proposes a view change to the cluster. Afterwards, it changes its
 \texttt{gen\_fsm} state to \lstinline!viewChange! and updates the
@@
 \lstinline!processBuffer! moves messages to the log, if all previous messages
-have now been received. The call to \lstinline!processLog! commits new messages
-in the log which have been committed by the master node. Lines 405 to 408
+have now been received. The call to \lstinline!processLog! commits logged
+messages which now have been committed by the master node. Lines 405 to 408
 destructure \lstinline!AfterLog! and bind variables for later use. The final
 \lstinline!if! statement responds to the master unless the master has committed
-messages we haven't received.\footnote{This can happen if the network drops a message whose
-  operation number lies between the replica's last commit number and the
-  master's last commit number.}
+messages we haven't received.\footnote{This can happen if the network drops a
+  message whose operation number lies between the replica's last commit number
+  and the master's last commit number.}
 
@@
 We distinguish between two types of failure: cessation of progress and loss of
-data. In a VR cluster without persistent stores, data loss only occurs when all
-$2f+1$ nodes in the cluster fail. In a VR cluster, progress only ceases if $f+1$
-nodes are lost. In \phatraid{} the conditions for cessation of progress and loss
-of data are complicated by the combination of RAID and VR.
+data. In a VR cluster without persistent stores, data loss of fully propagated,
+commited data only occurs when all $2f+1$ nodes in the cluster fail \footnote{If
+  more than $f+1$ nodes die some ``commited but not fully propogated data'' can
+  be lost}. In a VR cluster, progress only ceases if $f+1$ nodes are lost. In
+\phatraid{} the conditions for cessation of progress and loss of data are
+complicated by the combination of RAID and VR.
 
@@
 
-Four distributions of failed nodes in \phatraidcf{3}{1} are depicted in Figure
+Four arrangements of failed nodes in \phatraidcf{3}{1} are depicted in Figure
 \ref{fig:failures}. In this section, if we claim ``data is lost'', we mean that
 if no persistent store is used, data will be lost. Adding a persistent store
-obviously prevents loss of committed data.
+obviously prevents loss of committed data. Additionally, we assume node failures
+are independent and uniformly distributed throughout the \phatraid{} group.
 
-The top left corner depicts five of the nine nodes failing. This state is
-\emph{not} a failure state. The group can still make progress and no data is
-lost. The two surviving clusters can reach consensus on new values and can
-reconstruct the failed cluster's data.
+The top left corner, Figure \ref{fig:failures_allok}, depicts five of the nine
+nodes failing. This state is \emph{not} a failure state. The group can still
+make progress and no data is lost. The two surviving clusters can reach
+consensus on new values and can reconstruct the failed cluster's data.
 
-The top right corner depicts four of the nine nodes failing. This state is a
-failure state. The group cannot make progress but no data can been
-lost. Progress is blocked because more than one cluster cannot reach
-consensus. Since no cluster was completely destroyed, all data still exists on
-at least one node.
+The top right corner, Figure \ref{fig:failures_noprogress}, depicts four of the
+nine nodes failing. This state is a failure state. The group cannot make
+progress; however, no data can been lost. Progress is blocked because more than
+one cluster cannot reach consensus. Since no cluster was completely destroyed,
+all data still exists on at least one node.
 
-The bottom left corner is impossible because if data can be lost then we
-certainly cannot make progress.
+The bottom left corner, Figure \ref{fig:failures_impossible}, is impossible
+because if data can be lost then we certainly cannot make progress.
 
-The bottom right corner depicts six of the nine nodes failing. This state is
-also a failure state. The group cannot make progress and will lose
-data. Progress is blocked because more than one cluster cannot reach
-consensus. Additionally, two complete clusters have been lost so half of the
-data cannot be recovered.
-
-In general, progress ceases if more than one cluster cannot reach consensus and
-data loss occurs if more than one cluster completely fails. The lower bound of
-failed nodes for progress cessation in \phatraidcf{$C$}{$f$} is:
-    $$ 2\cdot(f+1) $$
-and the lower bound of failed nodes for data loss is:
-    $$ 2\cdot(2f+1) $$
-
-In contrast, if all data has been propagated to all nodes, data is only lost
-when all nodes die. If data has not been propagated to all nodes, then loss
-occurs if all nodes containing the data die.
+The bottom right corner, Figure \ref{fig:failures_noprogress_dataloss}, depicts
+six of the nine nodes failing. This state is also a failure state. The group
+cannot make progress and will lose data. Progress is blocked because more than
+one cluster cannot reach consensus. Additionally, two complete clusters have
+been lost so half of the data cannot be recovered.
 
@@
 
-A single number cannot describe the failure threshold of a \phatraid{} cluster
-because of the interplay between VR and RAID. The failure of the \phatraid{}
-group is dependent on the arrangement of the failed nodes. For example, Figure
-\ref{fig:failures_noprogress} depicts a situation wherein only four failures
-stops progress. In contrast, Figure \ref{fig:failures_allok} depicts a situation
-wherein five failures does \emph{not} stop progress.
+In general, progress ceases if more than one cluster \emph{cannot reach
+  consensus} and data loss occurs if more than one \emph{cluster completely
+  fails}. The lower bound of failed nodes for progress cessation in
+\phatraidcf{$C$}{$f$} is:
 
-In a \phatraidcf{3,1} group, less than one-quarter of four-failure arrangements
+    $$ 2\cdot(f+1) $$
+
+and the lower bound of failed nodes for data loss is:
+
+    $$ 2\cdot(2f+1) $$
+
+By the pidgeon hole principle, the upper bound of failed nodes which still
+permit progress is,
+
+    $$ C\cdot f + 1 $$
+
+because we can place $f$ failures in all but one cluster and place $f+1$
+failures in the final cluster.
+
+The upper bound for no data loss is,
+
+    $$ C\cdot 2f + 1 $$
+
+because we can keep one node of all but one cluster alive. From these $C-1$
+nodes we can reconstruct the entire file system.
+
+The bounds presented above cannot fully describe the failure threshold of a
+\phatraid{} cluster because of the interplay between VR and RAID. Instead, we
+will discuss the probability of group failure given $n$ node failures.
+
+Group failure is dependent on the arrangement of failed nodes. For example,
+Figure \ref{fig:failures_noprogress} depicts a situation wherein only four
+failures stops progress. In contrast, Figure \ref{fig:failures_allok} depicts a
+situation wherein five failures does \emph{not} stop progress.
+
+In a \phatraidcf{3}{1} group, less than one-quarter of four-failure arrangements
 stop progress. The necessary condition for progress cessation is that two or
@@
 
-When five nodes fail, they can be distributed into each group in the
-proportions: (3,1,1), (3,2,0) and (2,2,1). The first case does not stop progress
-because only one cluster fails. The later two cases stop progress because two
-clusters fail. We calculate the probability of \phatraid{} group failure when
-five nodes fail:
+When five nodes fail, they can be distributed into each group: (3,1,1), (3,2,0)
+and (2,2,1). The first case does not stop progress because only one cluster
+fails. The later two cases stop progress because two clusters fail. We calculate
+the probability of \phatraid{} group failure when five nodes fail:
 
@@
   \textrm{Failure Prob.}\;\;\;
-    &\Pr\left[\textrm{group failure} \,|\, n_{failed} = 5\right] = \frac{81 + 18}{126} =
-    \frac{99}{126} \approx 0.79 \\
+    &\Pr\left[\textrm{group failure} \,|\, \phatraidcf{3}{1},\, n_{failed} = 5\right] \\
+    &= \frac{81 + 18}{126} = \frac{99}{126} \approx 0.79 \\
 \end{align*}
@@
 1) - 1$ or fewer failures. A \phatraidcf{$C$}{$f$} group is likely resilient
-to $2(f+1)$ failures and unlikely resistant to $2(f+1)+1$ failures.
+to $2(f+1)$ failures and unlikely resistant to $C\cdot f + 1$ failures.
 
-The general form for the probability of \phatraidcf{$C$}{$f$} group failure
-given $2(f + 1)$ node failures is:
+For exactly $2(f+1)$ node failures, the probability of \phatraidcf{$C$}{$f$}
+group failure is:
 
 \begin{align*}
-  \Pr\left[\textrm{group failure} \,|\, \phatraidcf{C}{f}, n_{failure}=2(f+1)\right]
+  \Pr\left[\textrm{group failure} \,|\, \phatraidcf{C}{f},\, n_{failure}=2(f+1)\right]
     =
@@
 \begin{align*}
-  P\left[\textrm{group failure} \,|\, \phatraidcf{6}{1}, n_{failure}=4\right]
+  P\left[\textrm{group failure} \,|\, \phatraidcf{6}{1},\, n_{failure}=4\right]
     = \frac{135}{3000} \approx 0.05
@@
 
-Extra parity bits!
+We would like to consider adding more than one parity bit. For some
+applications, the diminishing returns on latency are not worth the loss of
+failures. In that case, we would do well to add more parity bits, thus allowing
+more clusters to completely fail.
 
 \section{Conclusion}
+
+We've presented RAID on a distributed file system which shows that sharding
+files improves latency for large files while maintaining reasonable failure
+probabilities.
 

### 2014-05-14 paper/paper.tex
Update paper

@@
 
+\begin{figure}
+  \includegraphics[scale=0.45]{Store.pdf}
+  \caption{Scatter Plot of Various \phatraidcf{$C$,$f$} Configurations}
+  \label{fig:plot}
+\end{figure}
+
+We evaluated a prototype \phatraidcf{C}{f} implementation. We simulated disk
+latency with sleeps. All experiments are run on a 2013-model MacBook Pro with
+8GB of RAM, and a 3GHz Intel Core i7. We found that latency scales as
+$\frac{1}{C-1}$. This is the expected behavior for breaking a file into $C-1$
+chunks. We always use one bit of parity.
+
+As expected \phatraidcf{2}{$f$} does not perform better than non-RAID \phat{}
+(i.e. \phatraidcf{1}{$f$}) because the data is not sharded into pieces, and the
+parity bit acts as a complete duplicate of the file.
+
 \section{Future Work}

### 2014-05-15 paper/paper.tex
more hd stuff

@@
 We believe that disk read/write latency is the major contributor to storing large files in a distributed system.
-{\bf TODO more}
 
@@
 \caption[Solid State Performance]{Solid State Disk performance for middle 95\% of solid state drives tested in a benchmark. Benchmark available at Benchmark available at \hwbenchmark.}
-\label{fig:sdd-perf}
+\label{fig:ssd-perf}
 \end{figure}
 
-Figure \ref{fig:sdd-perf} shows the performance for SSDs from a 2013 survey of commercially available SSDs.
+Figure \ref{fig:ssd-perf} shows the performance for SSDs from a 2013 survey of commercially available SSDs.
 Sequential write speeds measure the throughput (in megabytes per second) of data stored on the drive when the data is being written in the drive's natural sequential order (e.g.
@@
 We developed a novel ``{\bf {\em Disk Emulation and Abstraction Layer}}'' (DEAL)
-to simulate SSD performance so they .
+to simulate SSD performance for our evaluation. The DEAL simulates read and 
+write delay using the information from Figure \ref{fig:ssd-perf}. We calculate disk delay as:
+
+\begin{align}
+\textsc{ReadDelay}(x) =&\ x / \textsc{SeqReadSpeed} + \textsc{ReadAccessTime}\\
+\textsc{WriteDelay}(x) =&\ x / \textsc{SeqWriteSpeed} + \textsc{WriteAccessTime}
+\end{align}
+
+where $x$ is the size of the file. During filesystem operations, these formulas are
+used to put the process to sleep in proportion to the size of the file being written to
+or read from the drive.
+
+Instead of sending very large files, we sent smaller strings that were treated by the 
+DEAL as being much larger (e.g. 1 character represented 1MB of data). As a result,
+we extended the DEAL to perform an additional delay for performing bitwise XOR
+on file as that was the only operation on a file. In our model, drive delays were
+much greater than XOR delays (e.g. 20 GB/second XOR delay versus 200 MB/second
+drive delay).
 

### 2014-05-15 paper/paper.tex
future work

@@
 \begin{figure}
+\hspace*{-.43in}
   \centering
@@
 
+We would like to perform a real evaluation rather than a simulation. 
+It seems that as file size increases, PhatRaid becomes more appealing 
+for lower commit latency. It is not clear from our
+simulation, however, when the file size is small enough for other effects (such as
+network latency, communications overhead) to become more important. And, as a 
+result, PhatRaid having no improvements (or possibly worse performance).
+
 \section{Conclusion}

### 2014-05-15 paper/paper.tex
intro

@@
 new failure modes.
+
+Our setting is in a data center where we have low network latency,
+high network throughput, and the need to store large files with tolerance to
+some CPU, network, and disk failure. We reason that with these characteristics,
+disk read/write latency will be the major factor in overall latency to store a file
+with replication. As a result, we borrow ideas from RAID that allow for faster
+disk access, and mix them with ideas for distributed consensus algorithms that
+allow for consistent replication.
 

### 2014-05-15 paper/paper.tex
typos

@@
 In a review of hard disk performance \cite{disk-perf}, Hard Disk Drives (HDD) were compared with Solid State Drives (SSD).
-HDDs use a physical magnetic platter that spins so that a drive head can read or write data by sensing or manipulating the magnetic field in a particular position on to the platter.
+HDDs use a physical magnetic platter that spins so that a drive head can read or write data by sensing or manipulating the magnetic field in a particular position on the platter.
 In contrast, SSDs have no moving parts and instead uses integrated circuits as memory to store data persistently.
@@
 In contrast to HDDs, SSDs do not exhibit this problem as there is no spinning platter.
-The right diagram in \ref{fig:hdd-vs-sdd} shows relatively constant read performance throughout the entire disk.
+The left diagram in Figure \ref{fig:hdd-vs-sdd} shows relatively constant read performance throughout the entire disk.
 Additionally, HDDs have difficultly with random accesses as the drive may need to wait for the drive head to be directly over the correct position on the platter.
@@
 we extended the DEAL to perform an additional delay for performing bitwise XOR
-on file as that was the only operation on a file. In our model, drive delays were
-much greater than XOR delays (e.g. 20 GB/second XOR delay versus 200 MB/second
-drive delay).
+on a file as that was the only other operation on a file for our tests. 
+In our model, drive delays were much greater than XOR delays (e.g. 20 GB/second 
+XOR delay versus 200 MB/second drive delay).
 

### 2014-05-15 paper/paper.tex
future work
Update the paper

@@
 
-Our setting is in a data center where we have low network latency,
-high network throughput, and the need to store large files with tolerance to
-some CPU, network, and disk failure. We reason that with these characteristics,
-disk read/write latency will be the major factor in overall latency to store a file
-with replication. As a result, we borrow ideas from RAID that allow for faster
-disk access, and mix them with ideas for distributed consensus algorithms that
-allow for consistent replication.
+Our setting is in a data center where we have low network latency, high network
+throughput, and the need to store large files with tolerance to some CPU,
+network, and disk failure. In this setting, we reason that the major bottleneck
+for replicated file storage is disk read/write latency. We combat this
+bottleneck with ideas from RAID, which allow for faster disk access, and ideas
+from distributed consensus algorithms, which allow for consistent replication.
 
@@
 messages. Identifiers beginning with a lowercase letter are actually atoms, like
-Scheme's symbols. Identifiers beginning with an uppercase letter are normal
-identifiers.
+Scheme and Lisp's symbols. Identifiers beginning with an uppercase letter are
+normal identifiers.
 
@@
 maintain a connection to one of the $C$ \phat{} clusters. Each client process
-mediates all communication between its assigned \phat{} cluster and the
-raid-client.
+mediates communication between its assigned \phat{} cluster and the raid-client.
 
@@
 
-\subsection{Erlang Behaviors}
+\subsection{\phat{} Clusters}
 
-The server cluster is a collection of nodes running the same server code. One of
-these nodes is designated the master node (aka \phat{} Master). The other nodes are
-known as \phat{} replicas or just replicas.
+A server cluster is a collection of nodes which maintain a distributed file
+system. Each server cluster designates a unique master node (aka the \phat{}
+Master). The other nodes are known as \phat{} replicas or just replicas.
 
-The master and each replica has four components: the supervisor, the
-(unfortunately named) server, VR, and the file system:
+The master and each replica have four components: the supervisor, the server,
+VR, and the file system:
 
@@
 \end{itemize}
+
+\subsection{Erlang Behaviors}
 
@@
 
-Additionally, the \texttt{gen\_server} and \texttt{gen\_fsm} behaviors allow
-the programmer to specify an arbitrary state value which will be passed around
-\`{a} la functional reactive programming.
+Additionally, the \texttt{gen\_server} and \texttt{gen\_fsm} behaviors allow the
+programmer to specify an arbitrary state value which will be passed around \`{a}
+la functional reactive programming. We call this state value the \emph{store}.
 
@@
 Listing \ref{lst:supervisor} is taken from our supervisor code. When a
-supervisor is started, the \texttt{supervisor} behavior looks for a unary
-procedure called \lstinline!init!. The \lstinline!init! procedure returns a
-child specification which is a pair of a restart specification and a list of
+supervisor is started, the \texttt{supervisor} behavior looks for a procedure
+called \lstinline!init!. The \lstinline!init! procedure returns a child
+specification, which is a pair of a restart specification and a list of
 children. In Listing \ref{lst:supervisor}, the restart specification,
@@
 seconds. If more than one failure occurs in five seconds, the supervisor kills
-all the children and shuts itself down as well.
+all its children and then shuts itself down.
 
@@
 A state in the \texttt{gen\_fsm} behavior manifests as a procedure which accepts
-two arguments: the incoming message and the store\footnote{Mutable state is
-  implemented as a pure store value which is passed around by the behavior}. The
-return value of a state-procedure is usually the 4-tuple
+two arguments: the incoming message and the store. The return value of a
+state-procedure is usually the 4-tuple
 \lstinline!{next_state,NextState,NewStore,Timeout}!. The \texttt{gen\_fsm}
@@
 
-The \phat{} file system achieves consensus via the Viewstamped Replication
+The \phat{} file system reaches consensus via the Viewstamped Replication
 protocol \cite{liskov2012viewstamped}. Listing \ref{lst:genfsm} defines how to
@@
 \lstinline!prepareBuffer! stores prepare messages that arrived out of order. In
-particular, if the log ends at operation number $n$ and message $n+1$ is dropped
-by the network, all subsequent messages will be buffered. The buffered messages
-will not be added to the log until message $n+1$ is received. The call to
-\lstinline!processBuffer! moves messages to the log, if all previous messages
-have now been received. The call to \lstinline!processLog! commits logged
-messages which now have been committed by the master node. Lines 405 to 408
-destructure \lstinline!AfterLog! and bind variables for later use. The final
-\lstinline!if! statement responds to the master unless the master has committed
-messages we haven't received.\footnote{This can happen if the network drops a
-  message whose operation number lies between the replica's last commit number
-  and the master's last commit number.}
+particular, if the log ends at operation number $n-1$ and message $n$ is dropped
+by the network, all subsequent messages, $n+i$ will be buffered. The buffered
+messages will not be added to the log until message $n$ is received.
+
+The call to \lstinline!processBuffer! moves messages to the log, if all previous
+messages have now been received. The call to \lstinline!processLog! commits
+logged messages which have been newly committed by the master node. Lines
+405--408 destructure \lstinline!AfterLog! and bind variables for later use. The
+final \lstinline!if! statement responds to the master unless the master has
+committed messages we haven't yet received.\footnote{This can happen if the
+  network drops a message whose operation number lies between the replica's last
+  commit number and the master's last commit number.}
 
@@
 Four arrangements of failed nodes in \phatraidcf{3}{1} are depicted in Figure
-\ref{fig:failures}. In this section, if we claim ``data is lost'', we mean that
-if no persistent store is used, data will be lost. Adding a persistent store
-obviously prevents loss of committed data. Additionally, we assume node failures
-are independent and uniformly distributed throughout the \phatraid{} group.
+\ref{fig:failures}. In this section, ``data is lost'' means that, if no
+persistent, fail-safe store is used, some data will be lost. Adding a
+persistent, fail-safe store obviously prevents loss of committed
+data. Additionally, we assume node failures are independent and uniformly
+distributed throughout the \phatraid{} group.
 
@@
 nine nodes failing. This state is a failure state. The group cannot make
-progress; however, no data can been lost. Progress is blocked because more than
-one cluster cannot reach consensus. Since no cluster was completely destroyed,
-all data still exists on at least one node.
+progress; however, no data can been lost\footnote{Here and later in this paper
+  we mean that no data which has been decided and propagated throughout the
+  cluster can be lost. In VR, by contrast, if data is decided and propagated
+  throughout the cluster, that data can only be lost if every node is lost. We
+  also assume the nodes do not have persistent, fail-safe stores.}. Progress is
+blocked because more than one cluster cannot reach consensus. Since no cluster
+was completely destroyed, all data still exists on at least one node.
 
 The bottom left corner, Figure \ref{fig:failures_impossible}, is impossible
-because if data can be lost then we certainly cannot make progress.
+because the group certainly cannot make progress if data has been lost.
 
@@
 In general, progress ceases if more than one cluster \emph{cannot reach
-  consensus} and data loss occurs if more than one \emph{cluster completely
+  consensus} and data loss occurs if more than one cluster \emph{completely
   fails}. The lower bound of failed nodes for progress cessation in
@@
 By the pidgeon hole principle, the upper bound of failed nodes which still
-permit progress is,
+permit progress is:
 
-    $$ C\cdot f + 1 $$
+    $$ (C-1)\cdot f + 2f + 1 $$
 
-because we can place $f$ failures in all but one cluster and place $f+1$
-failures in the final cluster.
+because $C-1$ clusters can make progress, i.e. at least $f+1$ nodes are up in
+each cluster, and only one cluster is completely dead.
 
-The upper bound for no data loss is,
+The upper bound for no data loss is:
 
-    $$ C\cdot 2f + 1 $$
+    $$ (C-1)\cdot 2f + 2f + 1 $$
 
-because we can keep one node of all but one cluster alive. From these $C-1$
-nodes we can reconstruct the entire file system.
+because $C-1$ clusters have at least one node up and one cluster has no nodes
+up.
 
@@
 \phatraid{} cluster because of the interplay between VR and RAID. Instead, we
-will discuss the probability of group failure given $n$ node failures.
+discuss the probability of group failure given $n$ node failures.
 
@@
 
-When five nodes fail, they can be distributed into each group: (3,1,1), (3,2,0)
-and (2,2,1). The first case does not stop progress because only one cluster
-fails. The later two cases stop progress because two clusters fail. We calculate
-the probability of \phatraid{} group failure when five nodes fail:
+When five nodes fail, they can be distributed into each group in three different
+ways: (3,1,1), (3,2,0) and (2,2,1). The first case does not stop progress
+because only one cluster fails. The later two cases stop progress because two
+clusters fail. We calculate the probability of \phatraid{} group failure when
+five nodes fail:
 
@@
 1) - 1$ or fewer failures. A \phatraidcf{$C$}{$f$} group is likely resilient
-to $2(f+1)$ failures and unlikely resistant to $C\cdot f + 1$ failures.
+to $2(f+1)$ failures and unlikely resistant to $(C-1)\cdot f + 2f + 1$ failures.
 
-For exactly $2(f+1)$ node failures, the probability of \phatraidcf{$C$}{$f$}
+For exactly $2(f+1)$ node failures, the probability of a \phatraidcf{$C$}{$f$}
 group failure is:
@@
   \includegraphics[scale=0.45]{Store.pdf}
-  \caption{Scatter Plot of Various \phatraidcf{$C$,$f$} Configurations}
+  \caption{Scatter Plot of Various \phatraidcf{$C$}{$f$} Configurations}
   \label{fig:plot}
@@
 
-We evaluated a prototype \phatraidcf{C}{f} implementation. We simulated disk
+We evaluated a prototype \phatraidcf{$C$}{$f$} implementation. We simulated disk
 latency with sleeps. All experiments are run on a 2013-model MacBook Pro with
@@
 chunks. We always use one bit of parity.
+
+Figure \ref{fig:plot} is a scatter plot of the time to store a 100 MB file in a
+\phatraid{} group. Each point represents a different configuration. The x-axis
+indicates the number of nodes needed for that configuration. The x-axis can also
+be seen as a measure of cost to create the configuration.
+
+The slow linear growth from \phatraidcf{1}{1} to \phatraidcf{1}{14} reveals the
+cost of communication in large \phat{} clusters. This phenomenon is also present
+in the slight increase in cost from \phatraidcf{$C$}{1} to \phatraidcf{$C$}{2}.
 
@@
 simulation, however, when the file size is small enough for other effects (such as
-network latency, communications overhead) to become more important. And, as a 
-result, PhatRaid having no improvements (or possibly worse performance).
+network latency, communications overhead) to become more important.
 

### 2014-05-15 paper/paper.tex
Last updates

@@
 
-\title{PHAT RAID, yo}
+\title{\phatraid, yo}
 \author{Andrew Johnson \and Daniel King \and Lucas Waye \and Scott Moore}
@@
 
-Distributed file stores, like Google's \chubby{} and CS260r's \phat{}, send all
-files through a single master node which consistently replicates the file on
-many slaves. In this scheme, the master node's throughput is a bottleneck for
-storing large files. We present \phatraid{} which mitigates this bottleneck by
-partitioning files across many \paxos{} groups which form a \raid{1} group.
+Distributed file stores, like Google's \chubby{}\cite{burrows-chubby}
+and CS260r's \phat{}\cite{class}, send all files through a single
+master node which consistently replicates the file on many slaves. In
+this scheme, the master node's throughput is a bottleneck for storing
+large files. We present \phatraid{} which mitigates this bottleneck by
+partitioning files across many \paxos{} groups which form a \raid{}\cite{raid}
+group.
 
@@
 large files disk latency on the cluster nodes is a bottleneck. We propose
-building a RAID group out of distributed file system clusters. Large files are
+building a RAID (Redundant Array of Inexpensive Disks)\cite{raid} group out of distributed file system clusters. Large files are
 sharded and disk latency is reduced proportional to number of clusters. Failure
@@
 Our setting is in a data center where we have low network latency, high network
-throughput, and the need to store large files with tolerance to some CPU,
-network, and disk failure. In this setting, we reason that the major bottleneck
+throughput, and the need to dependably store large files with tolerance to some CPU,
+network, and disk failure. In this setting, we hypothesize that the major bottleneck
 for replicated file storage is disk read/write latency. We combat this
-bottleneck with ideas from RAID, which allow for faster disk access, and ideas
+bottleneck with ideas from RAID, which allow for faster file access, and ideas
 from distributed consensus algorithms, which allow for consistent replication.
@@
 
-Figure \ref{fig:hdd-vs-sdd} shows a comparison between read times for a SSD and HDD as a function of position on the disk.
-HDD storage performance is greatly influenced by where on the disk platter the data was placed.
-Data can be placed on the outer rim of the platter much more quickly than data placed near the center of the platter as the outer rim is moving faster in comparison (due to conservation of angular momentum).
-In contrast to HDDs, SSDs do not exhibit this problem as there is no spinning platter.
-The left diagram in Figure \ref{fig:hdd-vs-sdd} shows relatively constant read performance throughout the entire disk.
-Additionally, HDDs have difficultly with random accesses as the drive may need to wait for the drive head to be directly over the correct position on the platter.
-This problem becomes especially apparent when an individual file is fragmented into multiple blocks by the file system.
-SSDs have relatively better random access times, in general, than HDDs.
-Sequential access tends to be many times faster than random access for both SSDs and HDDs \cite{disk-perf}.
-HDD performance can also can suffer from {\em spin-up delay}, which happens as a result of a power-saving measure that stops spinning the disk during idle times.
+Figure \ref{fig:hdd-vs-sdd} shows a comparison between read times for
+a SSD and HDD as a function of position on the disk. HDD storage
+performance is greatly influenced by where on the disk platter the
+data was placed. Data can be placed on the outer rim of the platter
+much more quickly than data placed near the center of the platter as
+the outer rim is moving faster in comparison (due to conservation of
+angular momentum). In contrast to HDDs, SSDs do not exhibit this
+problem as there is no spinning platter. The left diagram in Figure
+\ref{fig:hdd-vs-sdd} shows relatively constant read performance
+throughout the entire disk. Additionally, HDDs have difficultly with
+random accesses as the drive may need to wait for the drive head to be
+directly over the correct position on the platter. This problem
+becomes especially apparent when an individual file is fragmented into
+multiple blocks by the file system. SSDs have better random access
+times, in general, than HDDs. Sequential access tends to be many times
+faster than random access for both SSDs and HDDs \cite{disk-perf}. HDD
+performance can also can suffer from {\em spin-up delay}, which
+happens as a result of a power-saving measure that stops spinning the
+disk during idle times.
 
-We decided that HDDs have far too much performance variability in order to be properly modeled.
-As a result, we instead chose to simulate the performance characteristics of SSDs as they have relatively constant access times regardless of sector position on the drive and also do not suffer from {\em spin-up delay}.
+We decided that HDDs have far too much performance variability in
+order to be properly modeled. As a result, we instead chose to
+simulate the performance characteristics of SSDs as they have
+relatively constant access times regardless of sector position on the
+drive and also do not suffer from {\em spin-up delay}. Also, because
+HDDs are slower, and our hypothesis supposes that disk is the
+bottleneck, this should be a more challenging case for \phatraid{}.
 
@@
 name. Erlang processes interact with one another by making RPC calls or sending
-messages. Identifiers beginning with a lowercase letter are actually atoms, like
+messages. Identifiers beginning with a lowercase letter are actually
+\emph{atoms}, a literal constant similar to
 Scheme and Lisp's symbols. Identifiers beginning with an uppercase letter are
@@
 \phat{}. The implementation is divided into three domains: the server cluster,
-the client, and the raid-client. The server code implements a VR-like consensus
-algorithm as well as simple file system. The client code exposes a file system
+the client, and the raid-client. The server code implements a variant
+of the Viewstamped Replication (VR)~\cite{liskov2012viewstamped} consensus
+algorithm as well a file system. The client code exposes a file system
 API which hides the communication bookkeeping. The raid-client provides a file
@@
 
-Each \phat{} cluster maintains a file system which is separate and distinct from
-the file systems of other \phat{} clusters. Generally, two phat clusters do not
-communicate. The raid-client and client processes all live on one node. Each
-\phat{} master and \phat{} replica lives on its own node.
+The master and each replica in a \phat{} cluster maintains a separate
+file system. Two different phat clusters do not communicate
+directly. The raid-client and client processes all live on one Erlang
+node. Each \phat{} master and \phat{} replica lives on its own Erlang
+node along with its file system.
 
@@
 In our model, drive delays were much greater than XOR delays (e.g. 20 GB/second 
-XOR delay versus 200 MB/second drive delay).
+XOR throughput versus 200 MB/second drive throughput).
 
@@
 
-On a client node, the raid-client spawns $C$ client processes which each
-maintain a connection to one of the $C$ \phat{} clusters. Each client process
+On a raid-client node, $C$ client processes are spawned each of which
+maintains a connection to one of the $C$ \phat{} clusters. Each client process
 mediates communication between its assigned \phat{} cluster and the raid-client.
@@
 \begin{itemize}
-\item the supervisor -- restarts the other three processes if any single node dies
+\item the supervisor -- restarts the other three processes if any dies
 \item the server -- passes messages to the VR process if this node is the
@@
   nodes in the cluster via the Viewstamped Replication protocol
-\item the file system -- maintains a simple hierarchical file system with locks
+\item the file system -- maintains a hierarchical file system with locks
 \end{itemize}
@@
 The four components are implemented using Erlang behaviors. An Erlang behavior
-is a framework which implements common patterns like servers and finite state
+is a framework which implements common patterns like servers and finite-state
 machines. In particular, the supervisor uses the \texttt{supervisor} behavior,
@@
   low-level socket interaction, messaging queueing, etc.
-\item The \texttt{gen\_fsm} behavior generalizes the \texttt{gen\_server}
-  behavior by permitting the server to have a finite number of states. Each
-  state has a set of functions with which to process messages. This behavior can
-  be seen as an ad hoc polymorphic variant of the \texttt{gen\_server} behavior.
+\item The \texttt{gen\_fsm} (finite-state machine) behavior
+generalizes the \texttt{gen\_server} behavior by permitting the server
+to have a finite number of states. Each state has a set of functions
+with which to process messages. This behavior can be seen as an ad hoc
+polymorphic variant of the \texttt{gen\_server} behavior.
 \end{itemize}
@@
 use the \lstinline!when! keyword to specify arbitrary required relationships
-between the matched variables.
+between the matched variables.\footnote{\lstinline{?debugFmt} is a macro
+    for printing debug messages when a flag is set}
 
@@
 final \lstinline!if! statement responds to the master unless the master has
-committed messages we haven't yet received.\footnote{This can happen if the
+committed messages the replica has not yet received.\footnote{This can happen if the
   network drops a message whose operation number lies between the replica's last
@@
 We distinguish between two types of failure: cessation of progress and loss of
-data. In a VR cluster without persistent stores, data loss of fully propagated,
+data. In a VR cluster data loss of fully propagated,
 commited data only occurs when all $2f+1$ nodes in the cluster fail \footnote{If
@@
 
-  \caption{Some Failed Node Distributions of \phatraidcf{3}{1}}
+  \caption{Some Failed Node Distributions of \phatraidcf{3}{1}. Each
+    circle represent a singe VR master or replica and each column is a
+  single \phat{} cluster. The red lines indicate failures.}
   \label{fig:failures}
@@
 
-Four arrangements of failed nodes in \phatraidcf{3}{1} are depicted in Figure
-\ref{fig:failures}. In this section, ``data is lost'' means that, if no
-persistent, fail-safe store is used, some data will be lost. Adding a
-persistent, fail-safe store obviously prevents loss of committed
-data. Additionally, we assume node failures are independent and uniformly
-distributed throughout the \phatraid{} group.
+Four arrangements of node failures in \phatraidcf{3}{1} are depicted in
+Figure \ref{fig:failures}. There are three columns, each representing
+a \phat cluster of three nodes. The red lines indicate failures. We
+assume failures are independent and uniformly distributed
+throughout the \phatraid{} group.
 
@@
 make progress and no data is lost. The two surviving clusters can reach
-consensus on new values and can reconstruct the failed cluster's data.
+consensus on new values and can reconstruct the failed cluster's data
+(by taking the XOR of the data on the two surviving clusters).
 
@@
 nine nodes failing. This state is a failure state. The group cannot make
-progress; however, no data can been lost\footnote{Here and later in this paper
+progress; however, no data can been lost.\footnote{Here and later in this paper
   we mean that no data which has been decided and propagated throughout the
-  cluster can be lost. In VR, by contrast, if data is decided and propagated
-  throughout the cluster, that data can only be lost if every node is lost. We
-  also assume the nodes do not have persistent, fail-safe stores.}. Progress is
-blocked because more than one cluster cannot reach consensus. Since no cluster
+  cluster can be lost. In VR and normal \phat, by contrast, if data is decided and propagated
+  throughout the cluster, that data can only be lost if every node is lost.} Progress is
+blocked because more than one cluster is unable to reach consensus. Since no cluster
 was completely destroyed, all data still exists on at least one node.
@@
 The bottom left corner, Figure \ref{fig:failures_impossible}, is impossible
-because the group certainly cannot make progress if data has been lost.
+because the group cannot make progress if data has been lost.
 
@@
 cannot make progress and will lose data. Progress is blocked because more than
-one cluster cannot reach consensus. Additionally, two complete clusters have
-been lost so half of the data cannot be recovered.
+one cluster is unable to reach consensus. Additionally, two complete clusters have
+been lost so at least half of the data cannot be recovered.\footnote{If
+  only the cluster with the parity bit remains then all useable data
+  has been lost}
 
@@
 
-By the pidgeon hole principle, the upper bound of failed nodes which still
+The upper bound of failed nodes which still
 permit progress is:
@@
 
-because $C-1$ clusters can make progress, i.e. at least $f+1$ nodes are up in
+because $C-1$ clusters could still make progress, e.g. if at least $f+1$ nodes are up in
 each cluster, and only one cluster is completely dead.
@@
 
-because $C-1$ clusters have at least one node up and one cluster has no nodes
-up.
+because it is possible that $C-1$ clusters have at least one node up
+and only one cluster has no nodes up.
 
@@
 stop progress. The necessary condition for progress cessation is that two or
-more clusters have lost at least $f+1$ nodes. By the pidgeon hole principle,
-only the two-in-two, four-failure arrangement stops progress. We can count the
+more clusters have lost at least $f+1$ nodes.
+Only the two-in-two, four-failure arrangement stops progress. We can count the
 number of such arrangements:
@@
 
-When five nodes fail, they can be distributed into each group in three different
+When five nodes fail, they can be distributed into each cluster in three different
 ways: (3,1,1), (3,2,0) and (2,2,1). The first case does not stop progress
@@
 
-The general formula for the failure probability distribution is not clear, but
+The general formula for the failure probability distribution is left
+as an exercise for the reader, but
 it is clear that a \phatraidcf{$C$}{$f$} group is absolutely resilient to $2(f +
-1) - 1$ or fewer failures. A \phatraidcf{$C$}{$f$} group is likely resilient
+1) - 1$ or fewer failures. In addition, a \phatraidcf{$C$}{$f$} group is likely resilient
 to $2(f+1)$ failures and unlikely resistant to $(C-1)\cdot f + 2f + 1$ failures.
@@
 
-For \phatraidcf{6}{1} we calculate:
+For example, for a \phatraidcf{6}{1} we calculate:
 
@@
 \begin{figure}
-  \includegraphics[scale=0.45]{Store.pdf}
+  \centering
+  \includegraphics[scale=0.5]{Store.png}
   \caption{Scatter Plot of Various \phatraidcf{$C$}{$f$} Configurations}
@@
 latency with sleeps. All experiments are run on a 2013-model MacBook Pro with
-8GB of RAM, and a 3GHz Intel Core i7. We found that latency scales as
+8GB of RAM, and a 3GHz Intel Core i7. We found that latency scales asp
 $\frac{1}{C-1}$. This is the expected behavior for breaking a file into $C-1$
-chunks. We always use one bit of parity.
+chunks. We always use one bit of party.
 
@@
 
-We would like to perform a real evaluation rather than a simulation. 
-It seems that as file size increases, PhatRaid becomes more appealing 
-for lower commit latency. It is not clear from our
-simulation, however, when the file size is small enough for other effects (such as
-network latency, communications overhead) to become more important.
+We would like to perform a real evaluation rather than a simulation.  It seems
+that as file size increases, PhatRaid becomes more appealing for lower commit
+latency. It is not clear from our simulation, however, when the file size is
+small enough for other effects (such as network latency, communications
+overhead) to become more important.
 
@@
 We've presented RAID on a distributed file system which shows that sharding
-files improves latency for large files while maintaining reasonable failure
+files can improve latency for large files while maintaining reasonable failure
 probabilities.

### 2014-05-16 paper/paper.tex
Some wording mods and bug fixes for the paper.

@@
 large files. We present \phatraid{} which mitigates this bottleneck by
-partitioning files across many \paxos{} groups which form a \raid{}\cite{raid}
+partitioning files across many \paxos{} groups which form a \raid{1}
 group.
@@
 large files disk latency on the cluster nodes is a bottleneck. We propose
-building a RAID (Redundant Array of Inexpensive Disks)\cite{raid} group out of distributed file system clusters. Large files are
+building a RAID group out of distributed file system clusters. Large files are
 sharded and disk latency is reduced proportional to number of clusters. Failure
@@
 \begin{figure}
-  \centering
-  \includegraphics[scale=0.5]{Store.png}
+  \includegraphics[scale=0.45]{Store.pdf}
   \caption{Scatter Plot of Various \phatraidcf{$C$}{$f$} Configurations}
@@
 latency with sleeps. All experiments are run on a 2013-model MacBook Pro with
-8GB of RAM, and a 3GHz Intel Core i7. We found that latency scales asp
+8GB of RAM, and a 3GHz Intel Core i7. We found that latency scales as
 $\frac{1}{C-1}$. This is the expected behavior for breaking a file into $C-1$
@@
 
-We would like to perform a real evaluation rather than a simulation.  It seems
-that as file size increases, PhatRaid becomes more appealing for lower commit
-latency. It is not clear from our simulation, however, when the file size is
-small enough for other effects (such as network latency, communications
-overhead) to become more important.
+We would like to perform a real evaluation rather than a simulation. 
+It seems that as file size increases, PhatRaid becomes more appealing 
+for lower commit latency. It is not clear from our
+simulation, however, when the file size is small enough for other effects (such as
+network latency, communications overhead) to become more important.
 
